
```{r load_libraries}
library(flowCore)
library(ggcyto)
```

## File Structure

A `flowFrame` is a single .FCS file exported from the flow cytometer's acquisition software. Multiple `flowFrame`s make up a `flowSet`, which represents all samples within a single experiment.

Here, we load in a single `flowFrame` and view a summary of the data. We will not transform the data here so we can see how it looks untransformed. Set `alter.names` to `TRUE` so that all parameter values will be more "R friendly."

```{r load_data}
file.name <- system.file("extdata", "0877408774.B08", package = "flowCore")

x <- read.FCS(file.name, transformation = FALSE, alter.names = TRUE)
summary(x)
```

## Stages of Flow Cytometry Data

It is important to understand where in the processing pipeline your data sits before doing anything in R. There are three distinct stages:

**Stage 1: Truly Raw**
The signal recorded across all detectors for every event — raw detector voltages. On a spectral instrument like the Cytek Aurora, this means one value per detector per cell (64 detectors on the 3-laser Aurora). You rarely work with this directly; it is what the acquisition software uses internally to perform unmixing.

**Stage 2: Unmixed** (Cytek Aurora specific)
SpectroFlo decomposes the full detector spectrum into contributions from each fluorochrome using your single stain reference spectra. The result is one value per fluorochrome per cell. This is what gets written to the unmixed FCS file and what you load into R. **Unmixed data is not raw** — a substantial mathematical operation has already been performed before you open R.

**Stage 3: Transformed**
You apply arcsinh or logicle in R to make the data visually and analytically interpretable. This is not done automatically, you decide what transformation to apply and do it yourself.

> **Reproducibility note:** Because unmixing happens before you ever open R, the quality of your entire downstream analysis is already determined by the quality of your single stain reference spectra, whether autofluorescence was properly extracted, and the unmixing settings in SpectroFlo. Errors at the unmixing stage cannot be corrected in R. You are always working with the consequences of those upstream decisions and this is true whether you analyze your flow data in FlowJo or in R.

## Checking the `$PnE` Keyword Before Transforming

Before applying any transformation, always check the `$PnE` keyword to confirm what you are actually working with. This keyword is part of the FCS file format specification and tells you whether the instrument already applied a transformation when it wrote the file.

```{r check_pne}
keyword(x, c("$P1E", "$P2E", "$P3E", "$P4E"))
```

`$PnE` values of `"0,0"` indicate untransformed linear data. Anything other than `"0,0"` means a transformation was already applied using the following formula:

$$y = f_2 \times 10^{\frac{f_1 \cdot x}{R}}$$

where $x$ is the original channel value, $y$ is the scaled value, $f_1$ and $f_2$ are given respectively by the first and second element of the `$PnE` key, and $R$ is the channel range given by the value of the `$PnR` key.

> **Reproducibility note:** Some acquisition software exports transformed FCS files while others export raw linear values. If different lab members export files differently from the same instrument, your analyses will not be comparable even though they came from the same experiment. Always verify `$PnE` on any new dataset.

## Common Data Transformations in Flow Cytometry

Raw flow data has two problems that make it difficult to work with directly:

1. **The range is enormous.** Highly expressed markers might have 100x more signal than dimly expressed ones. On a linear scale, dim populations get crushed into an unresolvable blob near zero.
2. **Negative values exist.** After compensation or unmixing, cells with truly no expression of a marker can end up with slightly negative values due to noise. A standard log transform cannot handle negatives, which is why simple log transformation does not work for flow data.

### Logicle (Biexponential)
The standard transformation for conventional flow cytometry. Linear near zero — which handles negatives gracefully — and logarithmic at high values. See Parks et al., 2006 in *Cytometry A* for the canonical reference.

### Arcsinh
Popularized by mass cytometry (CyTOF) and now standard for most modern high-dimensional flow analysis, including spectral flow. Behaves similarly to logicle but is mathematically simpler. **This is the preferred transformation for Cytek Aurora data.** The cofactor parameter controls where the linear-to-log transition occurs (typically 150 for flow cytometry). See Bendall et al., 2011 in *Science*.

### Applying Arcsinh Transformation in R

```{r apply_transform}
# Define the arcsinh transformation with cofactor of 150
asinhTrans <- arcsinhTransform(transformationId = "arcsinh", a = 0, b = 1/150, c = 0)

# Apply to fluorochrome parameters (adjust column indices to match your data)
transList <- transformList(colnames(x)[4:ncol(x)], asinhTrans)
x_transformed <- transform(x, transList)
```

> **Note:** Transformation is applied to fluorochrome parameters only, not scatter parameters (FSC, SSC), which are already on a meaningful linear scale.

### Loading Multiple Samples into a Single `flowSet`
If we have a directory containing FCS files (most experiments will have multiple samples) we can read in a list of those files and create a flowSet out of them:

```{r}
fcs.dir <- system.file("extdata", "compdata", "data", package="flowCore")
frames <- lapply(dir(fcs.dir, full.names=TRUE), read.FCS)
fs <- as(frames, "flowSet")
fs

sampleNames(fs)
```

Note that the original list is unnamed and that the resulting sample names are not particularly meaningful. If the list is named, the list constructed is much more meaningful. One such approach is to employ the `keyword` method for `flowFrame` objects to extract the `“SAMPLE ID”` keyword from each frame:

```{r}
names(frames) <- sapply(frames, keyword, "SAMPLE ID")
fs <- as(frames, "flowSet")
fs

sampleNames(fs)
```