---
title: "01: Load Data and Attach Metadata"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    includes:
      after_body: _nav.html
---

<div style="display:flex; gap:8px; margin-bottom:1.5em;">
<a href="02_qc.html" class="btn btn-primary btn-sm">Next: QC →</a>
</div>

---

```{r libraries, message=FALSE, warning=FALSE}
library(flowCore)
library(flowSpecs)
library(tidyverse)
```

## Setting up a working directory

Before running anything, set a consistent working directory. Scripts 02 onwards save processed files and each script must find files created by the previous one.

```r
dir.create("~/flow_tutorial", showWarnings = FALSE)
setwd("~/flow_tutorial")
```

You only need to create the folder once. Each time you restart R, run `setwd("~/flow_tutorial")` before continuing.

---

## Data structures

All flow cytometry data in R lives in two objects from the `flowCore` package:

**`flowFrame`** — a single FCS file. Contains a matrix of expression values (one row per cell event and one column per measured parameter), parameter metadata, and FCS file keywords. Access the expression data directly with `exprs()`.

**`flowSet`** — a collection of `flowFrame`s from one experiment. Lets you apply operations across all samples at once via `fsApply()`. Most of your work in this tutorial happens at the `flowSet` level.

## Stages of spectral flow data

Before loading anything, it is important to understand where in the processing pipeline your data sits.

**Stage 1: Truly raw** — raw detector voltages off the instrument. On the Cytek Aurora, the instrument records signal across all detectors simultaneously for every cell. This is what we work with in this tutorial.

**Stage 2: Unmixed** — spectral unmixing decomposes the full detector spectrum into per-fluorochrome values using single-stain reference spectra. Normally this happens in SpectroFlo before you open R. In this tutorial we demonstrate unmixing in R using a pre-computed unmixing matrix, which is exactly what SpectroFlo does under the hood.

**Stage 3: Transformed** — you apply arcsinh in R to compress the range and handle negative values from unmixing. This is your responsibility, covered in script 03.

> **Why this matters:** Errors at the unmixing stage — poor single-stains, missing autofluorescence extraction, inconsistent reference control gating — cannot be corrected in R. You are always working with the consequences of those upstream decisions.

---

## Loading the example data

This tutorial uses the `fullPanel` dataset from the `flowSpecs` Bioconductor package — a human PBMC sample measured on a spectral flow cytometer. The data is exactly as it comes off the instrument: raw detector intensities, untransformed, requiring spectral unmixing before any biological interpretation is possible.

```{r load}
data(fullPanel)   # raw spectral PBMC flowFrame
data(unmixCtrls)  # single-stain unmixing controls (15 samples)
data(specMat)     # pre-computed spectral unmixing matrix

fullPanel
```

This is a single sample with 8,000 cells. A real experiment would have multiple samples; the pipeline mechanics demonstrated here apply equally to any number.

## Understanding the channel structure

Unlike conventional or pre-unmixed spectral flow data, the channels here are named by detector position rather than by fluorochrome or marker:

```{r channels}
colnames(fullPanel)
```

Each group of channels corresponds to one laser:

| Channels | Laser | Excitation |
|---|---|---|
| `V1-A` – `V16-A` | Violet | 405 nm |
| `B1-A` – `B16-A` | Blue | 488 nm |
| `R1-A` – `R10-A` | Red | 640 nm |

The `A` suffix denotes area measurements. `FSC-A`, `FSC-H`, `SSC-A`, `SSC-H` are scatter parameters as usual.

> **On your own Aurora data:** Your unmixed FCS files from SpectroFlo will have fluorochrome-named channels (e.g., `BV421-A`, `FITC-A`) rather than detector positions. Load them with:
> ```r
> fcs_dir <- "data/unmixed"
> fcs_files <- list.files(fcs_dir, pattern = "\\.fcs$", full.names = TRUE)
> frames <- lapply(fcs_files, read.FCS, transformation = FALSE, alter.names = TRUE)
> names(frames) <- basename(fcs_files)
> fs <- as(frames, "flowSet")
> ```
> Since SpectroFlo has already performed unmixing, you would skip script 03's unmixing step and go straight to transformation.

## The panel

The unmixing matrix `specMat` tells us which fluorochromes — and therefore which markers — are in this panel:

```{r panel}
panel <- data.frame(
  fluorochrome = rownames(specMat),
  marker = c("IgM", "CD4", "CD19", "CD14", "CD56", "CD11c",
             "CD8a", "CD41b", "CD3", "Unknown", "CD45RA",
             "Viability", "Autofluorescence"),
  class = c("type", "type", "type", "type", "type", "type",
            "type", "type", "type", "none", "type", "QC", "none")
)
panel
```

This PBMC panel covers the major immune populations: T cells (CD3, CD4, CD8a, CD45RA), B cells (CD19, IgM), NK cells (CD56), monocytes (CD14), dendritic cells (CD11c), and platelets (CD41b), with a viability dye to exclude dead cells.

## Wrapping into a flowSet

The rest of the pipeline expects a `flowSet`. With a single sample, wrapping it is one line:

```{r to_flowset}
fs <- flowSet(list(PBMC = fullPanel))
sampleNames(fs)
nrow(fs[[1]])
```

> **For multi-sample experiments**, attach metadata to `pData()` after building the flowSet:
> ```r
> sample_meta     <- read_csv("metadata/sample_metadata.csv")
> experiment_meta <- read_csv("metadata/experiment_metadata.csv")
> meta <- left_join(sample_meta, experiment_meta, by = "experiment_id") |> as.data.frame()
> meta_ordered <- meta[match(sampleNames(fs), meta$filename), ]
> rownames(meta_ordered) <- meta_ordered$filename
> pData(fs) <- meta_ordered
> ```
> Always reorder metadata rows to match `sampleNames(fs)` exactly before assigning.

The next script runs QC on this raw spectral data before unmixing occurs.

<hr>
<div style="display:flex; gap:8px; margin-top:1.5em;">
<a href="02_qc.html" class="btn btn-primary btn-sm">Next: QC →</a>
</div>
