---
title: "01: Load Data and Attach Metadata"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
---

<div style="display:flex; gap:8px; margin-bottom:1.5em;">
<a href="../index.html" class="btn btn-default btn-sm">← Home</a>
<a href="02_qc.html" class="btn btn-primary btn-sm">Next: QC →</a>
</div>

---

## Project structure and file naming

### Directory structure

Here is an example of how I suggest you structure your project directory.

```
data/
  raw/              Unprocessed FCS files directly from instrument — never modified
  unmixed/          Unmixed FCS files exported from SpectroFlo — starting point for R analysis
  processed/        Intermediate objects generated by the scripts (gitignored)
metadata/
  experiment_metadata.csv
  sample_metadata.csv
scripts/
  01_load_and_metadata.Rmd  Data structures, stages of flow data, load files, attach metadata
  02_qc.Rmd                 QC: temporal stability, margin events
  03_transformation.Rmd     Check $PnE, apply arcsinh
  04_gating.Rmd             Sequential manual gating
  05_dimensionality_reduction.Rmd  UMAP
  06_clustering.Rmd         FlowSOM (optional)
  07_analysis.Rmd           Frequencies, MFI, statistical comparisons
output/
  figures/
  tables/
```

### FCS file naming

Use underscores as separators and avoid spaces and special characters. Spaces in filenames cause problems when loading files programmatically. A consistent naming scheme also makes files easier to sort and identify at a glance. A useful convention:

```
YYYYMMDD_mouseID_sex_treatment_timepoint_sampleSource_panel.fcs
```

For example: `20240603_M01_F_Salmonella_D7_SPL_CD8tet.fcs`

Define any abbreviations you plan to use (sample sources, panel names, treatment groups) in your experiment metadata before data collection begins, and don't change them mid-project.

### Setting up a working directory

Scripts 02 onwards save processed files, and each script must be able to find files created by the previous one. Set a working directory once and use the same folder throughout the tutorial.

```r
# Create a folder for this tutorial and set it as your working directory.
# You can change this path to wherever you like.
dir.create("~/flow_tutorial", showWarnings = FALSE)
setwd("~/flow_tutorial")
```

You only need to create the folder once. Each time you restart R, run `setwd("~/flow_tutorial")` before continuing.

---

## Stages of flow data

Before loading anything, it is important to understand where in the processing pipeline your data sits.

**Stage 1: Truly raw** — raw detector voltages off the instrument. On a spectral instrument like the Cytek Aurora, one value per detector per cell across all 64 detectors. You never work with this directly.

**Stage 2: Unmixed** (Aurora-specific) — SpectroFlo decomposes the full detector spectrum into per-fluorochrome values using single-stain reference spectra. **This is what gets written to the unmixed FCS file and what you load into R.** A substantial mathematical step has already occurred before you open R.

**Stage 3: Transformed** — you apply arcsinh or logicle in R to compress the range and handle negative values. This is your responsibility, covered in script 03.

> **Why this matters:** Errors at the unmixing stage — poor single-stains, missing autofluorescence extraction, inconsistent reference control gating — cannot be corrected in R. You are always working with the consequences of those upstream decisions.

---

## Loading the example data

```{r libraries, message=FALSE, warning=FALSE}
library(flowCore)
library(tidyverse)
```

This tutorial uses the `GvHD` dataset built into the `flowCore` package which is a longitudinal study of bone marrow transplant patients. Five patients were each measured at seven visits, giving 35 samples total. The panel is a conventional 4-color FITC/PE/PerCP/APC setup run on a FACScan. Differences from spectral data are noted throughout.

## Data structures

All flow cytometry data in R lives in two objects from the `flowCore` package:

**`flowFrame`** — a single FCS file. Contains a matrix of expression values (one row per event and one column per measured parameter), parameter metadata, and FCS file keywords. Access the expression data directly with `exprs()`.

**`flowSet`** — a collection of `flowFrame`s from one experiment. Lets you apply operations across all samples at once via `fsApply()`. Most of your work in this tutorial happens at the `flowSet` level.

Use the code below to load your data and explore the `flowSet`:
```{r load_gvhd}
data(GvHD)
GvHD
```

The column names in this dataset use dashes (e.g., `FL1-H`). R works better with dots, so we normalize them with `make.names()`. This is exactly what `alter.names = TRUE` does when loading FCS files with `read.FCS()`.

```{r normalize_colnames}
colnames(GvHD) <- make.names(colnames(GvHD))
colnames(GvHD)
```

> **On your own data:** You would load FCS files from disk instead:
> ```r
> fcs_dir <- "data/unmixed"
> fcs_files <- list.files(fcs_dir, pattern = "\\.fcs$", full.names = TRUE)
> frames <- lapply(fcs_files, read.FCS, transformation = FALSE, alter.names = TRUE)
> names(frames) <- basename(fcs_files)
> fs <- as(frames, "flowSet")
> ```
> Always use `transformation = FALSE` so R loads raw linear values rather than applying its own default transformation. You will apply the transformation yourself in script 03.
>
> **Spectral flow note:** Unmixed FCS files will have one column per fluorochrome rather than one per detector. The number of columns depends on your panel size, not the number of detectors.

## Inspecting the flowSet

```{r inspect}
sampleNames(GvHD)
```

Each sample name encodes patient and visit: `s5a01` = patient 5, visit 1. Use `nrow()` on a flowFrame to see the event count for each sample.

```{r event_counts}
fsApply(GvHD, nrow)
```

## Viewing and adding sample metadata

Every `flowSet` carries per-sample metadata in a data frame accessible via `pData()`. The GvHD dataset already includes useful fields:

```{r pdata}
pData(GvHD)
```

| Column | Meaning |
|--------|---------|
| `Patient` | Patient ID |
| `Visit` | Visit number (1–7) |
| `Days` | Actual days post-transplant (negative = pre-transplant baseline) |
| `Grade` | GvHD severity grade at this visit (1 = mild, 3 = severe) |

We add three columns that downstream scripts expect:

```{r add_metadata}
meta <- pData(GvHD)
meta$patient_id <- as.character(meta$Patient)
meta$treatment <- ifelse(meta$Grade == 1, "Grade 1 (mild)", "Grade 3 (severe)")
meta$timepoint_days <- meta$Days
pData(GvHD) <- meta
pData(GvHD)[, c("patient_id", "treatment", "timepoint_days")]
```

`treatment` is derived from GvHD severity grade, a real biological variable. Patients 5, 6, 7, and 9 have Grade 3 (severe GvHD); patient 10 has Grade 1 (mild). With only five patients this is a small dataset, but it is large enough to demonstrate the full pipeline.

> **On your own data:** Your metadata would come from CSV files rather than a built-in dataset:
> ```r
> sample_meta <- read_csv("metadata/sample_metadata.csv")
> experiment_meta <- read_csv("metadata/experiment_metadata.csv")
> meta <- left_join(sample_meta, experiment_meta, by = "experiment_id") |> as.data.frame()
> meta_ordered <- meta[match(sampleNames(fs), meta$filename), ]
> rownames(meta_ordered) <- meta_ordered$filename
> pData(fs) <- meta_ordered
> ```
> Always reorder metadata rows to match `sampleNames(fs)` exactly before assigning — `pData()` does not reorder automatically.

The next script runs QC on this data. No save is needed here — GvHD is small enough that each script reloads it directly.

<hr>
<div style="display:flex; gap:8px; margin-top:1.5em;">
<a href="../index.html" class="btn btn-default btn-sm">← Home</a>
<a href="02_qc.html" class="btn btn-primary btn-sm">Next: QC →</a>
</div>
