---
title: "01: Load Data and Attach Metadata"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
---

> **Navigation:** [Home](../index.html) | [Next: QC →](02_qc.html)

---

```{r libraries, message=FALSE, warning=FALSE}
library(flowCore)
library(tidyverse)
```

We start by reading the unmixed FCS files from `data/unmixed/`. We name each frame by its filename immediately so that when we build the flowSet, `sampleNames()` returns something meaningful rather than the default numeric index. This naming matters because we'll use the filename as the join key when we attach metadata.

```{r load_fcs}
fcs_dir <- "data/unmixed"
fcs_files <- list.files(fcs_dir, pattern = "\\.fcs$", full.names = TRUE)

frames <- lapply(fcs_files, read.FCS, transformation = FALSE, alter.names = TRUE)
names(frames) <- basename(fcs_files)

fs <- as(frames, "flowSet")
fs
```

Load the two metadata files separately, then join them on `experiment_id`. Keeping them separate until this point makes it easier to validate each one before merging.

```{r load_metadata, message=FALSE}
sample_meta <- read_csv("metadata/sample_metadata.csv")
experiment_meta <- read_csv("metadata/experiment_metadata.csv")

meta <- left_join(sample_meta, experiment_meta, by = "experiment_id") |> as.data.frame()
```

Before attaching metadata to the flowSet, reorder its rows to match `sampleNames(fs)` exactly. `pData()` expects a one-to-one correspondence and does not reorder automatically.

```{r attach_metadata}
meta_ordered <- meta[match(sampleNames(fs), meta$filename), ]
rownames(meta_ordered) <- meta_ordered$filename

pData(fs) <- meta_ordered
pData(fs)
```

A quick sanity check: the number of rows in `pData(fs)` should equal the number of samples, every filename should match, and no metadata fields should be NA for samples that should have values.

```{r sanity_check}
stopifnot(nrow(pData(fs)) == length(sampleNames(fs)))
stopifnot(all(rownames(pData(fs)) == sampleNames(fs)))
summary(is.na(pData(fs)))
```

The metadata is now attached. The next script re-loads from the original FCS files and applies QC filtering — no intermediate save is needed since the source files are already in `data/unmixed/` and the metadata is already in `metadata/`.

---

> **Navigation:** [Home](../index.html) | [Next: QC →](02_qc.html)
