---
title: "02: Quality Control"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
---

<div style="display:flex; gap:8px; margin-bottom:1.5em;">
<a href="../index.html" class="btn btn-default btn-sm">← Home</a>
<a href="01_load_and_metadata.html" class="btn btn-default btn-sm">← Prev: Load Data</a>
<a href="03_transformation.html" class="btn btn-primary btn-sm">Next: Transformation →</a>
</div>

---

```{r libraries, message=FALSE, warning=FALSE}
library(flowCore)
library(flowWorkspace)
library(PeacoQC)
library(tidyverse)
```

```{r load, message=FALSE}
data(GvHD)
colnames(GvHD) <- make.names(colnames(GvHD))
meta <- pData(GvHD)
meta$patient_id     <- as.character(meta$Patient)
meta$treatment      <- ifelse(meta$Grade == 1, "Grade 1 (mild)", "Grade 3 (severe)")
meta$timepoint_days <- meta$Days
pData(GvHD) <- meta
fs <- GvHD
```

Before transforming or gating anything, we check for acquisition problems. The two main things to look for are signal instability over time (caused by clogging or fluidics issues mid-run) and margin events (events where the signal hit the detector ceiling and the true value is unknown).

Plotting signal against time is the fastest way to spot a run that went wrong. A healthy acquisition looks flat. A sudden spike, gradual drift, or a section of flat line followed by a jump usually means something happened to the flow rate.

```{r time_plots, fig.width=12, fig.height=16, warning=FALSE, message=FALSE}
time_data <- fsApply(fs, function(ff) {
  data.frame(
    Time  = exprs(ff)[, "Time"],
    FL1.H = exprs(ff)[, "FL1.H"],
    sample = identifier(ff)
  )
}, simplify = FALSE)
time_df <- do.call(rbind, time_data)

ggplot(time_df, aes(x = Time, y = FL1.H)) +
  geom_hex(bins = 40) +
  facet_wrap(~sample, ncol = 5) +
  scale_fill_viridis_c() +
  labs(title = "Signal stability over time — FL1.H channel",
       x = "Time", y = "FL1.H") +
  theme_minimal(base_size = 10)
```

PeacoQC automates this detection. It identifies and removes anomalous events based on signal consistency over time. We run it on each sample and collect the cleaned frames.

> **Spectral flow note:** On spectral data, run PeacoQC on a representative subset of unmixed channels rather than all of them. A channel with very few positive cells can produce noisy signal that PeacoQC flags as instability when it isn't. Choose channels that have enough positive events across most samples.

```{r peaco_qc, message=FALSE, warning=FALSE, results='hide'}
dir.create("output/qc", recursive = TRUE, showWarnings = FALSE)

fluor_channels <- c("FL1.H", "FL2.H", "FL3.H", "FL4.H")

fs_qc <- fsApply(fs, function(ff) {
  result <- PeacoQC(
    ff               = ff,
    channels         = fluor_channels,
    output_directory = "output/qc",
    name             = identifier(ff),
    plot             = FALSE,
    save_fcs         = FALSE
  )
  result$FinalFF
})
```

After running PeacoQC, check how many events were removed from each sample. Losing a small fraction is expected — anything over 20% is worth investigating further.

If a sample exceeds that threshold, go back to the time plots and look at *where* the instability occurred. A brief blip at the very start of acquisition is common and less concerning as the flow rate often takes a moment to stabilise as the sample first enters the system. However, instability spread throughout the run is more serious.

Also check whether the remaining cell count is still adequate. Losing 25% of 100,000 events is generally fine; losing 25% of 3,000 may leave you with too few cells for reliable gating and statistics.

- **If instability was brief and cell count is adequate:** keep the sample, flag it, and note the deviation in your experiment metadata.
- **If instability was severe or cell count is insufficient:** exclude the sample and document it in your methods. If cells are still viable, reacquisition is worth considering.

The most important principle is that exclusion decisions must be made and documented before looking at your results, regardless of what you choose to do.

```{r qc_summary}
events_before <- as.integer(fsApply(fs, nrow))
events_after  <- as.integer(fsApply(fs_qc, nrow))

data.frame(
  sample       = sampleNames(fs),
  before       = events_before,
  after        = events_after,
  pct_retained = round(events_after / events_before * 100, 1)
)
```

Margin events are events where the detector recorded a value at the maximum of its range, meaning the true signal was even higher and got clipped. As a result, we remove them because the actual value is unknown. This is distinct from what PeacoQC catches; PeacoQC looks at temporal instability, not ceiling effects.

```{r margin_events}
remove_margin_events <- function(ff) {
  channel_max <- parameters(ff)@data$maxRange
  names(channel_max) <- colnames(ff)
  at_ceiling <- apply(exprs(ff), 1, function(row) {
    any(row >= channel_max * 0.9999, na.rm = TRUE)
  })
  ff[!at_ceiling, ]
}

fs_clean <- fsApply(fs_qc, remove_margin_events)
```

```{r qc_report}
data.frame(
  sample          = sampleNames(fs_clean),
  final_events    = as.integer(fsApply(fs_clean, nrow)),
  pct_of_original = round(as.integer(fsApply(fs_clean, nrow)) /
                          as.integer(fsApply(fs, nrow)) * 100, 1)
)
```

We convert the QC-cleaned flowSet to a GatingSet, which is the standard persistence format in the `flowWorkspace` ecosystem. All downstream scripts load from this. Even with no gates applied yet, `save_gs()` correctly serializes the data and metadata.

```{r save}
dir.create("data/processed", recursive = TRUE, showWarnings = FALSE)
gs <- GatingSet(fs_clean)
unlink("data/processed/GvHD_gs_qc", recursive = TRUE)
save_gs(gs, path = "data/processed/GvHD_gs_qc")
cat("QC-cleaned GatingSet saved to data/processed/GvHD_gs_qc\n")
```

<hr>
<div style="display:flex; gap:8px; margin-top:1.5em;">
<a href="../index.html" class="btn btn-default btn-sm">← Home</a>
<a href="01_load_and_metadata.html" class="btn btn-default btn-sm">← Prev: Load Data</a>
<a href="03_transformation.html" class="btn btn-primary btn-sm">Next: Transformation →</a>
</div>
