---
title: "02: Quality Control"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
---

> **Navigation:** [Home](../index.html) | [← Prev: Load Data](01_load_and_metadata.html) | [Next: Transformation →](03_transformation.html)

---

```{r libraries, message=FALSE, warning=FALSE}
library(flowCore)
library(PeacoQC)
library(tidyverse)
```

```{r load, message=FALSE}
fcs_files <- list.files("data/unmixed", pattern = "fcs$", full.names = TRUE)
frames    <- lapply(fcs_files, read.FCS, transformation = FALSE, alter.names = TRUE)
names(frames) <- basename(fcs_files)
fs <- as(frames, "flowSet")

sample_meta     <- read_csv("metadata/sample_metadata.csv", show_col_types = FALSE)
experiment_meta <- read_csv("metadata/experiment_metadata.csv", show_col_types = FALSE)
meta <- left_join(sample_meta, experiment_meta, by = "experiment_id") |> as.data.frame()
meta_ordered <- meta[match(sampleNames(fs), meta$filename), ]
rownames(meta_ordered) <- meta_ordered$filename
pData(fs) <- meta_ordered
```

Before transforming or gating anything, we check for acquisition problems. The two main things to look for are signal instability over time (caused by clogging or fluidics issues mid-run) and margin events (events where the signal hit the detector ceiling and the true value is unknown).

Plotting signal against time is the fastest way to spot a run that went wrong. A healthy acquisition looks flat. A sudden spike, gradual drift, or a section of flat line followed by a jump usually means something happened to the flow rate.

```{r time_plots, fig.width=14, fig.height=8, warning=FALSE, message=FALSE}
time_data <- fsApply(fs, function(ff) {
  data.frame(
    Time  = exprs(ff)[, "Time"],
    FL1.H = exprs(ff)[, "FL1.H"],
    sample = identifier(ff)
  )
}, simplify = FALSE)
time_df <- do.call(rbind, time_data)

ggplot(time_df, aes(x = Time, y = FL1.H)) +
  geom_hex(bins = 40) +
  facet_wrap(~sample, ncol = 7) +
  scale_fill_viridis_c() +
  labs(title = "Signal stability over time — FL1.H channel",
       x = "Time", y = "FL1.H") +
  theme_minimal(base_size = 8)
```

PeacoQC automates this detection. It identifies and removes anomalous events based on signal consistency over time. We run it on each sample and collect the cleaned frames.

```{r peaco_qc, message=FALSE, warning=FALSE, results='hide'}
dir.create("output/qc", recursive = TRUE, showWarnings = FALSE)

fluor_channels <- c("FL1.H", "FL2.H", "FL3.H", "FL4.H")

fs_qc <- fsApply(fs, function(ff) {
  result <- PeacoQC(
    ff               = ff,
    channels         = fluor_channels,
    output_directory = "output/qc",
    name             = identifier(ff),
    plot             = FALSE,
    save_fcs         = FALSE
  )
  result$FinalFF
})
```

After running PeacoQC, check how many events were removed from each sample. Losing a small fraction is expected — anything over 20% is worth going back to look at the time plots for that sample.

```{r qc_summary}
events_before <- as.integer(fsApply(fs, nrow))
events_after  <- as.integer(fsApply(fs_qc, nrow))

data.frame(
  sample       = sampleNames(fs),
  before       = events_before,
  after        = events_after,
  pct_retained = round(events_after / events_before * 100, 1)
)
```

Margin events are events where the detector recorded a value at the maximum of its range — the true signal was even higher and got clipped. We remove them because the actual value is unknown. This is distinct from what PeacoQC catches; PeacoQC looks at temporal instability, not ceiling effects.

```{r margin_events}
remove_margin_events <- function(ff) {
  channel_max <- parameters(ff)@data$maxRange
  names(channel_max) <- colnames(ff)
  at_ceiling <- apply(exprs(ff), 1, function(row) {
    any(row >= channel_max * 0.9999, na.rm = TRUE)
  })
  ff[!at_ceiling, ]
}

fs_clean <- fsApply(fs_qc, remove_margin_events)
```

```{r qc_report}
data.frame(
  sample          = sampleNames(fs_clean),
  final_events    = as.integer(fsApply(fs_clean, nrow)),
  pct_of_original = round(as.integer(fsApply(fs_clean, nrow)) /
                          as.integer(fsApply(fs, nrow)) * 100, 1)
)
```

We save each QC-passed frame as an individual FCS file. The `flowSet` class uses reference-based storage that doesn't serialize cleanly with `saveRDS`, so saving individual frames with `write.FCS()` is the correct approach at this stage.

```{r save}
dir.create("data/processed/qc", recursive = TRUE, showWarnings = FALSE)
fsApply(fs_clean, function(ff) {
  write.FCS(ff, file.path("data/processed/qc", identifier(ff)))
})
cat("Saved", length(fs_clean), "QC-filtered FCS files to data/processed/qc/\n")
```

---

> **Navigation:** [Home](../index.html) | [← Prev: Load Data](01_load_and_metadata.html) | [Next: Transformation →](03_transformation.html)
