---
title: "03: Transformation"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
---

<div style="display:flex; gap:8px; margin-bottom:1.5em;">
<a href="../index.html" class="btn btn-default btn-sm">← Home</a>
<a href="02_qc.html" class="btn btn-default btn-sm">← Prev: QC</a>
<a href="04_gating.html" class="btn btn-primary btn-sm">Next: Gating →</a>
</div>

---

```{r libraries, message=FALSE, warning=FALSE}
library(flowCore)
library(flowWorkspace)
library(ggcyto)
library(tidyverse)
```

```{r load, message=FALSE}
qc_files <- list.files("data/processed/qc", pattern = "fcs$", full.names = TRUE)
frames <- lapply(qc_files, read.FCS, transformation = FALSE, alter.names = TRUE)
names(frames) <- basename(qc_files)
fs <- as(frames, "flowSet")

meta <- read.csv("data/processed/qc/GvHD_metadata.csv", row.names = 1)
pData(fs) <- meta[sampleNames(fs), ]
```

## Why transformation is necessary

Raw flow cytometry data has two properties that make it difficult to work with directly:

1. **The range is enormous.** Highly expressed markers can have 100× more signal than dimly expressed ones. On a linear scale, dim populations collapse into an unresolvable cluster near zero.
2. **Negative values exist.** After compensation or spectral unmixing, cells with no true expression of a marker can have slightly negative values due to detector noise. Standard log transformation breaks on negatives.

Two transformations address both problems. Which one you use depends on how your data was collected:

**Logicle (biexponential)** — the standard for conventional flow cytometry. Linear near zero (handles negatives gracefully) and logarithmic at high values. See Parks et al., 2006 in *Cytometry A*.

**Arcsinh** — the standard for mass cytometry (CyTOF) and modern spectral flow including the Cytek Aurora. The `b` parameter controls the linear-to-log transition (`b = 1/cofactor`; cofactor 150 is standard for flow cytometry). Apply only to fluorochrome channels — scatter parameters (FSC, SSC) are already on a meaningful linear scale and should not be transformed. See Bendall et al., 2011 in *Science*.

The GvHD dataset was collected on a conventional cytometer with hardware log amplifiers, which means the transformation was applied before the signal was even digitized. This makes the `$PnE` check especially important here.

## Checking `$PnE` before transforming

Before applying any transformation, check the `$PnE` keyword to confirm how the data is stored. `$PnE = "0,0"` means untransformed linear — you apply arcsinh yourself. Anything else means a transformation was already applied at acquisition or export.

```{r check_pne}
keyword(fs[[1]], paste0("$P", 1:7, "E"))
```

For this dataset, the scatter channels (FSC.H, SSC.H) are `"0,0"` — linear, as expected. The fluorescence channels (FL1.H, FL2.H, FL3.H, FL4.H) are `"4,0"`, which means the data was collected through a log amplifier and stored on a 4-decade log scale. **This data is already transformed.** Applying arcsinh on top of it would be wrong and would make the data difficult to interpret.

This is a fundamental difference between conventional cytometers and modern digital instruments like the Cytek Aurora. On the Aurora, data is captured as linear detector voltages and stored with `$PnE = "0,0"` — you choose and apply the transformation yourself in R. On older conventional cytometers with analog electronics, the log transformation happens in hardware before the signal is digitized. Always check `$PnE` before doing anything.

For reference, the `"4,0"` format means the stored channel value $x$ was generated from the original signal $y$ as:

$$x = \frac{R}{f_1} \log_{10}(y)$$

where $R$ is the channel range and $f_1 = 4$ is the number of log decades. So the raw stored value is already on a log scale with 4 decades of dynamic range.

Since this data is pre-transformed, we skip arcsinh and visualize the fluorescence channels directly. You should see clear populations.

```{r visualize, fig.width=12, fig.height=6, warning=FALSE, message=FALSE}
ggcyto(fs[1:6], aes(x = "FL1.H", y = "FL2.H")) +
  geom_hex(bins = 50) +
  facet_wrap(~name) +
  labs(title = "FL1.H vs FL2.H — log-amplified, no additional transformation needed")
```

We convert to a GatingSet at this step, which is how `flowWorkspace` persists the data for all downstream gating, UMAP, and analysis scripts. Even with no gates applied yet, `save_gs()` correctly serializes the flowSet data. When working with Aurora data where transformation is applied, the transformed flowSet would be converted here.

```{r save}
gs <- GatingSet(fs)
unlink("data/processed/GvHD_tcell_gs_base", recursive = TRUE)
save_gs(gs, path = "data/processed/GvHD_tcell_gs_base")
cat("GatingSet saved to data/processed/GvHD_tcell_gs_base\n")
```

> For Cytek Aurora data: load the QC-cleaned flowSet, verify `$PnE = "0,0"` for all fluorochrome channels, then apply arcsinh with cofactor 150:
> ```r
> fluor_channels <- c("B515.A", "V525.A", ...)  # your panel's channels
> asinhTrans <- arcsinhTransform(transformationId = "arcsinh", a = 0, b = 1/150, c = 0)
> transList <- transformList(fluor_channels, asinhTrans)
> fs_transformed <- transform(fs, transList)
> ```

<hr>
<div style="display:flex; gap:8px; margin-top:1.5em;">
<a href="../index.html" class="btn btn-default btn-sm">← Home</a>
<a href="02_qc.html" class="btn btn-default btn-sm">← Prev: QC</a>
<a href="04_gating.html" class="btn btn-primary btn-sm">Next: Gating →</a>
</div>
