---
title: "03: Transformation"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    includes:
      after_body: _nav.html
---

<div style="display:flex; gap:8px; margin-bottom:1.5em;">
<a href="02_qc.html" class="btn btn-default btn-sm">← Prev: QC</a>
<a href="04_gating.html" class="btn btn-primary btn-sm">Next: Gating →</a>
</div>

---

```{r libraries, message=FALSE, warning=FALSE}
library(flowCore)
library(flowWorkspace)
library(ggcyto)
library(tidyverse)
```

```{r load, message=FALSE}
gs <- load_gs("data/processed/GvHD_gs_qc")
```

## Why transformation is necessary

Raw flow cytometry data has two properties that make it difficult to work with directly:

1. **The range is enormous.** Highly expressed markers can have 100× more signal than dimly expressed ones. On a linear scale, dim populations collapse into an unresolvable cluster near zero.
2. **Negative values exist.** After compensation or spectral unmixing, cells with no true expression of a marker can have slightly negative values due to detector noise. Standard log transformation breaks on negatives.

Two transformations address both problems. Which one you use depends on how your data was collected:

**Logicle (biexponential)** — the standard for conventional flow cytometry. Linear near zero (handles negatives gracefully) and logarithmic at high values. See Parks et al., 2006 in *Cytometry A*.

**Arcsinh** — the standard for mass cytometry (CyTOF) and modern spectral flow. The `b` parameter controls the linear-to-log transition (`b = 1/cofactor`; cofactor 150 is standard for flow cytometry). Apply only to fluorochrome channels: scatter parameters (FSC, SSC) are already on a meaningful linear scale and should not be transformed. See Bendall et al., 2011 in *Science*.

The GvHD dataset was collected on a conventional cytometer with hardware log amplifiers, which means the transformation was applied before the signal was even digitized. This makes the `$PnE` check especially important here.

## Checking `$PnE` before transforming

Run this on every new dataset before doing anything else:

```{r check_pne}
keyword(gs_pop_get_data(gs, "root")[[1]], paste0("$P", 1:7, "E"))
```

The scatter channels (FSC.H, SSC.H) are `"0,0"` — untransformed linear values, as expected. The fluorescence channels (CD15_FITC_H, CD45_PE_H, CD14_PerCP_H, CD33_APC_H) are `"4,0"`, which means a log transformation was already applied in hardware before the signal was digitized. **Applying another transformation on top would be wrong.**

On a modern instrument like the Cytek Aurora, fluorescence channels are stored as linear values (`$PnE = "0,0"`) and you apply the transformation yourself in R. On older conventional cytometers with analog electronics, the log transformation happens before digitization and arrives baked into the file. The `"4,0"` value tells you the data is on a 4-decade log scale:

$$x = \frac{R}{f_1} \log_{10}(y)$$

where $x$ is the stored channel value, $y$ is the original signal, $R$ is the channel range, and $f_1 = 4$ is the number of log decades.

Since this data is pre-transformed, we skip straight to visualization.

```{r visualize, fig.width=12, fig.height=6, warning=FALSE, message=FALSE}
ggcyto(gs[1:6], aes(x = "CD15_FITC_H", y = "CD45_PE_H"), subset = "root") +
  geom_hex(bins = 50) +
  facet_wrap(~name) +
  labs(title = "CD15 (FITC) vs CD45 (PE) — log-amplified, no additional transformation needed")
```

The GatingSet loaded from script 02 is passed through as-is — no transformation applied for this dataset. For Aurora data, you would apply arcsinh to the flowSet before this point and the transformed data would be carried forward here.

```{r save}
unlink("data/processed/GvHD_tcell_gs_base", recursive = TRUE)
save_gs(gs, path = "data/processed/GvHD_tcell_gs_base")
cat("GatingSet saved to data/processed/GvHD_tcell_gs_base\n")
```

> For spectral data: load the QC-cleaned flowSet, verify `$PnE = "0,0"` for all fluorochrome channels, then apply arcsinh with cofactor 150:
> ```r
> fluor_channels <- c("B515.A", "V525.A", ...)  # your panel's channels
> asinhTrans <- arcsinhTransform(transformationId = "arcsinh", a = 0, b = 1/150, c = 0)
> transList <- transformList(fluor_channels, asinhTrans)
> fs_transformed <- transform(fs, transList)
> ```

<hr>
<div style="display:flex; gap:8px; margin-top:1.5em;">
<a href="02_qc.html" class="btn btn-default btn-sm">← Prev: QC</a>
<a href="04_gating.html" class="btn btn-primary btn-sm">Next: Gating →</a>
</div>
