---
title: "04: Gating"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
---

<div style="display:flex; gap:8px; margin-bottom:1.5em;">
<a href="../index.html" class="btn btn-default btn-sm">← Home</a>
<a href="03_transformation.html" class="btn btn-default btn-sm">← Prev: Transformation</a>
<a href="05_dimensionality_reduction.html" class="btn btn-primary btn-sm">Next: Dimensionality Reduction →</a>
</div>

---

```{r libraries, message=FALSE, warning=FALSE}
library(flowCore)
library(flowWorkspace)
library(openCyto)
library(tidyverse)
```

```{r load}
gs <- load_gs("data/processed/GvHD_tcell_gs_base")
```

```{r plot_helper, include=FALSE}
# Helper: extract expression data from a gating population for ggplot2 plotting
get_pop_df <- function(gs, pop) {
  fsApply(gs_pop_get_data(gs, pop), function(ff) {
    as.data.frame(exprs(ff)) %>% mutate(sample = identifier(ff))
  }, simplify = FALSE) %>% do.call(rbind, .)
}
```

Gating is sequential — each gate is applied within the parent population from the previous step. The order here is: all events → remove debris → remove doublets → gate population of interest. Every downstream number in this analysis is a direct consequence of where these gates are drawn.

**A note on this dataset:** the GvHD data was collected on a conventional FACScan with only height (H) and one area (A) parameter — specifically FL2.A. Modern instruments record FSC-Area and SSC-Area which are the standard channels for singlet discrimination. We'll work with what this dataset provides, and differences from your own Aurora data are noted throughout.

Start with the scatter gate to isolate lymphocytes from debris. Visualize first to understand the data before setting values.

```{r scatter_vis, fig.width=14, fig.height=8, warning=FALSE}
df_all <- get_pop_df(gs, "root")

ggplot(df_all, aes(x = FSC.H, y = SSC.H)) +
  geom_hex(bins = 60) +
  scale_fill_viridis_c() +
  facet_wrap(~sample, ncol = 7) +
  theme_minimal(base_size = 8) +
  labs(title = "All events — identify lymphocyte cluster before setting gate")
```

```{r scatter_gate}
scatter_gate <- rectangleGate(
  filterId = "lymphocytes",
  "FSC.H"  = c(100, 850),
  "SSC.H"  = c(0, 500)
)

gs_pop_add(gs, scatter_gate, parent = "root")
recompute(gs)
```

```{r scatter_check, fig.width=14, fig.height=8, warning=FALSE}
ggplot(df_all, aes(x = FSC.H, y = SSC.H)) +
  geom_hex(bins = 60) +
  scale_fill_viridis_c() +
  geom_rect(aes(xmin = 100, xmax = 850, ymin = 0, ymax = 500),
            fill = NA, color = "red", linewidth = 0.8, inherit.aes = FALSE) +
  facet_wrap(~sample, ncol = 7) +
  theme_minimal(base_size = 8) +
  labs(title = "Lymphocyte gate (red box)")
```

For singlet discrimination, modern instruments use FSC-Height vs FSC-Area: single cells fall on a tight diagonal, doublets appear above it. This dataset doesn't have FSC-Area, so we use FL2.H vs FL2.A — the PE channel height and area — which captures the same doublet-exclusion principle. On your Aurora data, use `FSC.A` vs `FSC.H` instead.

```{r singlet_vis, fig.width=14, fig.height=8, warning=FALSE}
df_lymph <- get_pop_df(gs, "lymphocytes")

ggplot(df_lymph, aes(x = FL2.A, y = FL2.H)) +
  geom_hex(bins = 60) +
  scale_fill_viridis_c() +
  facet_wrap(~sample, ncol = 7) +
  theme_minimal(base_size = 8) +
  labs(title = "FL2.A vs FL2.H — singlet population falls along the diagonal")
```

```{r singlet_gate}
singlet_gate <- rectangleGate(
  filterId = "singlets",
  "FL2.A"  = c(0, 900),
  "FL2.H"  = c(1, 5000)
)

gs_pop_add(gs, singlet_gate, parent = "lymphocytes")
recompute(gs)
```

The GvHD panel doesn't include a viability dye. In a real experiment you would gate out dead cells here using a viability dye channel — dead cells stain brightly, live cells are dim or negative. Since we can't demonstrate this with the GvHD data, we proceed directly to the population gate.

> For Aurora data, add a live/dead gate after singlets:
> ```r
> live_gate <- rectangleGate(filterId = "live", "V525.A" = c(-Inf, 2))
> gs_pop_add(gs, live_gate, parent = "singlets")
> recompute(gs)
> ```

Now gate on your population of interest. Here we gate on FL1.H-high cells (the FITC channel).

```{r population_vis, fig.width=14, fig.height=8, warning=FALSE}
df_sing <- get_pop_df(gs, "singlets")

ggplot(df_sing, aes(x = FL1.H, y = FL4.H)) +
  geom_hex(bins = 60) +
  scale_fill_viridis_c() +
  facet_wrap(~sample, ncol = 7) +
  theme_minimal(base_size = 8) +
  labs(title = "FL1.H vs FL4.H — gate on your population of interest")
```

```{r population_gate}
population_gate <- rectangleGate(
  filterId         = "population_of_interest",
  "FL1.H"          = c(1000, 10000),
  "FL4.H"          = c(1, 10000)
)

gs_pop_add(gs, population_gate, parent = "singlets")
recompute(gs)
```

Check the cell counts at each gate across all samples. Large drops between steps are worth investigating.

```{r gate_stats}
gs_pop_get_stats(gs, nodes = gs_get_pop_paths(gs), type = "count")
```

```{r save}
save_gs(gs, path = "data/processed/GvHD_tcell_gs")
cat("Gated GatingSet saved\n")
```

<hr>
<div style="display:flex; gap:8px; margin-top:1.5em;">
<a href="../index.html" class="btn btn-default btn-sm">← Home</a>
<a href="03_transformation.html" class="btn btn-default btn-sm">← Prev: Transformation</a>
<a href="05_dimensionality_reduction.html" class="btn btn-primary btn-sm">Next: Dimensionality Reduction →</a>
</div>
