---
title: "05: Dimensionality Reduction"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
---

<div style="display:flex; gap:8px; margin-bottom:1.5em; align-items:center; flex-wrap:wrap;">
<a href="../index.html" class="btn btn-default btn-sm">← Home</a>
<div class="btn-group">
<button type="button" class="btn btn-default btn-sm dropdown-toggle" data-toggle="dropdown">Scripts <span class="caret"></span></button>
<ul class="dropdown-menu">
<li><a href="01_load_and_metadata.html">01: Load and Metadata</a></li>
<li><a href="02_qc.html">02: QC</a></li>
<li><a href="03_transformation.html">03: Transformation</a></li>
<li><a href="04_gating.html">04: Gating</a></li>
<li><a href="05_dimensionality_reduction.html">05: Dimensionality Reduction</a></li>
<li><a href="06_clustering.html">06: Clustering</a></li>
<li><a href="07_analysis.html">07: Analysis</a></li>
</ul>
</div>
<a href="04_gating.html" class="btn btn-default btn-sm">← Prev: Gating</a>
<a href="06_clustering.html" class="btn btn-primary btn-sm">Next: Clustering →</a>
</div>

---

```{r libraries, message=FALSE, warning=FALSE}
library(flowWorkspace)
library(uwot)
library(ggplot2)
library(tidyverse)
```

```{r load}
gs <- load_gs("data/processed/GvHD_tcell_gs")
```

UMAP takes all your fluorochrome channels and compresses them into two dimensions, letting you visualize the overall structure of your data — which populations exist, how distinct they are, and whether samples cluster by treatment or by something unexpected (like batch or operator). This is an exploratory step, not an analytical one. What you see here should guide interpretation, not replace gating or statistics.

We run UMAP on the singlet-gated cells and subsample before running — UMAP doesn't need every cell to find the structure, and running it on very large datasets is slow with minimal benefit.

```{r extract_expression}
fluor_channels <- c("CD15_FITC_H", "CD45_PE_H", "CD14_PerCP_H", "CD33_APC_H")

singlet_data <- gs_pop_get_data(gs, "singlets")

expr_list <- fsApply(singlet_data, function(ff) {
  as.data.frame(exprs(ff)[, fluor_channels]) %>%
    mutate(sample = identifier(ff))
}, simplify = FALSE)

expr_df <- bind_rows(expr_list)
```

```{r subsample}
set.seed(42)
n_per_sample <- 500  # small dataset so this keeps it fast

expr_sub <- expr_df %>%
  group_by(sample) %>%
  slice_sample(n = n_per_sample) %>%
  ungroup()
```

```{r run_umap, message=FALSE}
umap_coords <- umap(
  expr_sub[, fluor_channels],
  n_neighbors = 15,
  min_dist    = 0.3,
  n_threads   = 4,
  verbose     = FALSE
)

umap_df <- expr_sub %>%
  mutate(UMAP1 = umap_coords[, 1], UMAP2 = umap_coords[, 2])
```

Plot colored by each marker to understand what's driving the structure. Regions with high expression of a marker should correspond to populations you'd recognize from gating.

```{r marker_plots, fig.width=10, fig.height=8, warning=FALSE}
plot_marker <- function(df, channel) {
  ggplot(df, aes(x = UMAP1, y = UMAP2, color = .data[[channel]])) +
    geom_point(size = 0.5, alpha = 0.6) +
    scale_color_viridis_c() +
    labs(title = channel) +
    theme_minimal() +
    theme(legend.position = "right")
}

p1 <- plot_marker(umap_df, "CD15_FITC_H")
p2 <- plot_marker(umap_df, "CD45_PE_H")
p3 <- plot_marker(umap_df, "CD14_PerCP_H")
p4 <- plot_marker(umap_df, "CD33_APC_H")

gridExtra::grid.arrange(p1, p2, p3, p4, ncol = 2)
```

Also plot colored by metadata to check whether samples are mixing properly or separating by something unexpected. Samples from the same treatment group should largely overlap. A clear separation by patient or visit rather than by disease grade could indicate a batch effect worth investigating before statistics.

```{r metadata_plots, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
meta <- pData(gs) %>% rownames_to_column("sample")
umap_df <- left_join(umap_df, meta, by = "sample")

p_treatment <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = treatment)) +
  geom_point(size = 0.5, alpha = 0.5) +
  theme_minimal() +
  labs(title = "Colored by treatment")

p_patient <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = patient_id)) +
  geom_point(size = 0.5, alpha = 0.5) +
  theme_minimal() +
  labs(title = "Colored by patient — check for patient-level batch effects")

gridExtra::grid.arrange(p_treatment, p_patient, ncol = 2)
```

```{r save}
saveRDS(umap_df, "data/processed/GvHD_tcell_umap.rds")
```

<hr>
<div style="display:flex; gap:8px; margin-top:1.5em; align-items:center; flex-wrap:wrap;">
<a href="../index.html" class="btn btn-default btn-sm">← Home</a>
<div class="btn-group">
<button type="button" class="btn btn-default btn-sm dropdown-toggle" data-toggle="dropdown">Scripts <span class="caret"></span></button>
<ul class="dropdown-menu">
<li><a href="01_load_and_metadata.html">01: Load and Metadata</a></li>
<li><a href="02_qc.html">02: QC</a></li>
<li><a href="03_transformation.html">03: Transformation</a></li>
<li><a href="04_gating.html">04: Gating</a></li>
<li><a href="05_dimensionality_reduction.html">05: Dimensionality Reduction</a></li>
<li><a href="06_clustering.html">06: Clustering</a></li>
<li><a href="07_analysis.html">07: Analysis</a></li>
</ul>
</div>
<a href="04_gating.html" class="btn btn-default btn-sm">← Prev: Gating</a>
<a href="06_clustering.html" class="btn btn-primary btn-sm">Next: Clustering →</a>
</div>
