```{r libraries}
library(flowWorkspace)
library(FlowSOM)
library(tidyverse)
library(ggplot2)
```

```{r load}
gs <- load_gs("data/processed/20240603_CD8tet_gs")
umap_df <- readRDS("data/processed/20240603_CD8tet_umap.rds")
```

This script is for discovery mode — when you want to find populations without defining them manually through gating. If your panel has predefined populations you already know how to gate (e.g. CD8 T cells, NK cells), you may not need this step. Where clustering becomes useful is for high-dimensional panels where manual gating can't capture the full complexity of a population (e.g. identifying functional subsets within your gated population of interest, or doing an unbiased survey of all live cells).

FlowSOM works by first building a self-organizing map (SOM) on your expression data, then grouping SOM nodes into metaclusters using hierarchical clustering. The two parameters you control are `xdim`/`ydim` (the size of the SOM grid, which determines resolution) and `maxMeta` (the maximum number of metaclusters to try). Start with `maxMeta = 15` and adjust based on what you see — too few clusters merge populations that should be distinct, too many over-split.

```{r extract_for_clustering}
live_data <- gs_pop_get_data(gs, "live")
fluor_channels <- colnames(live_data[[1]])[4:ncol(live_data[[1]])]

# Subsample for clustering — 2000 cells per sample is usually enough to capture structure
set.seed(42)
expr_list <- fsApply(live_data, function(ff) {
  idx <- sample(nrow(ff), min(2000, nrow(ff)))
  df <- as.data.frame(exprs(ff)[idx, fluor_channels])
  df$sample <- identifier(ff)
  df
}, simplify = FALSE)

expr_df <- bind_rows(expr_list)
```

```{r run_flowsom}
fsom <- FlowSOM(
  input = as.matrix(expr_df[, fluor_channels]),
  xdim = 10,
  ydim = 10,
  maxMeta = 15,
  seed = 42
)
```

Add cluster assignments back to the expression dataframe, then join with UMAP coordinates so we can visualize which UMAP regions correspond to which clusters.

```{r add_clusters}
expr_df$cluster <- as.factor(GetMetaclusters(fsom))

# Join with UMAP — they share the same subsampled cells if the seed was identical
umap_clustered <- umap_df %>%
  mutate(cluster = expr_df$cluster)  # this assumes the row order is the same — verify if you change the subsampling

ggplot(umap_clustered, aes(x = UMAP1, y = UMAP2, color = cluster)) +
  geom_point(size = 0.3, alpha = 0.5) +
  theme_minimal() +
  labs(title = "FlowSOM metaclusters on UMAP")
```

To understand what each cluster represents, look at the median expression of each marker per cluster. This is the equivalent of a marker heatmap in scRNAseq — clusters should have distinct marker combinations that correspond to known populations.

```{r cluster_heatmap}
cluster_medians <- expr_df %>%
  group_by(cluster) %>%
  summarise(across(all_of(fluor_channels), median)) %>%
  pivot_longer(-cluster, names_to = "channel", values_to = "median_expression")

ggplot(cluster_medians, aes(x = channel, y = cluster, fill = median_expression)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Median marker expression per FlowSOM cluster")
```

Annotate clusters once you've identified what each one is. Some clusters may require splitting further or merging if they don't correspond to a biologically meaningful distinction.

```{r annotate_clusters}
# Fill in cluster labels based on your marker heatmap interpretation
cluster_labels <- c(
  "1" = "CD8 Effector",
  "2" = "CD8 Memory",
  "3" = "CD4 T cell"
  # ... add the rest
)

expr_df$population <- cluster_labels[as.character(expr_df$cluster)]
```

```{r save}
saveRDS(expr_df, "data/processed/20240603_CD8tet_clustered.rds")
```
