---
title: "06: Clustering (Optional)"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    includes:
      after_body: _nav.html
---

<div style="display:flex; gap:8px; margin-bottom:1.5em;">
<a href="05_dimensionality_reduction.html" class="btn btn-default btn-sm">← Prev: Dimensionality Reduction</a>
<a href="07_analysis.html" class="btn btn-primary btn-sm">Next: Analysis →</a>
</div>

---

```{r libraries, message=FALSE, warning=FALSE}
library(flowWorkspace)
library(FlowSOM)
library(tidyverse)
library(ggplot2)
```

```{r load}
gs      <- load_gs("data/processed/PBMC_gs_gated")
umap_df <- readRDS("data/processed/PBMC_umap.rds")
umap_df <- umap_df[, !colnames(umap_df) %in% "Original_ID"]
```

This script is for discovery mode — when you want to find populations without defining them manually through gating. For PBMC data with a panel covering all major leukocyte lineages, unsupervised clustering can identify the major populations automatically and confirm (or challenge) your gating strategy. It is also useful for identifying rare populations or functional subsets that might not be obvious from gating alone.

FlowSOM builds a self-organizing map on your expression data, then groups SOM nodes into metaclusters using hierarchical clustering. Start conservatively with `maxMeta` and increase if clusters look biologically merged.

```{r extract}
marker_channels <- c("CD3_PB", "CD4_AF700", "CD8a_BV785",
                     "CD14_BV605", "CD19_APCCy7", "CD56_BV650",
                     "CD11c_BV711", "CD45RA_PECy7", "IgM_AF647")

live_data <- gs_pop_get_data(gs, "live")
expr_df   <- as.data.frame(exprs(live_data[[1]])[, marker_channels])
```

```{r run_flowsom}
set.seed(42)
fsom <- FlowSOM(
  input   = as.matrix(expr_df),
  xdim    = 8,
  ydim    = 8,
  maxMeta = 8,
  seed    = 42
)
```

Overlay cluster assignments on the UMAP. Each cluster should correspond to a region of the UMAP — if a cluster spans multiple disconnected regions it is probably splitting a real population or merging two distinct ones.

```{r umap_clusters, fig.width=6, fig.height=5, warning=FALSE}
expr_df$cluster <- as.factor(GetMetaclusters(fsom))

# The subsampled rows must match the umap_df rows (same seed)
umap_clustered <- umap_df %>%
  mutate(cluster = expr_df$cluster)

ggplot(umap_clustered, aes(x = UMAP1, y = UMAP2, color = cluster)) +
  geom_point(size = 0.5, alpha = 0.6) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal(base_size = 11) +
  labs(title = "FlowSOM metaclusters on UMAP")
```

The cluster heatmap shows median expression of each marker per cluster. Use this to label what each cluster represents biologically:

```{r heatmap, fig.width=9, fig.height=4, warning=FALSE}
cluster_medians <- expr_df %>%
  group_by(cluster) %>%
  summarise(across(all_of(marker_channels), median)) %>%
  pivot_longer(-cluster, names_to = "channel", values_to = "median_expr")

ggplot(cluster_medians, aes(x = channel, y = cluster, fill = median_expr)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Median marker expression per FlowSOM cluster",
       x = NULL, y = "Cluster", fill = "Median\n(arcsinh)")
```

Expected cluster identities for a PBMC panel:
- **CD3+ CD4+** → CD4 T cells (helper)
- **CD3+ CD8a+** → CD8 T cells (cytotoxic)
- **CD3+ CD4+ CD45RA+** → naive CD4 T cells
- **CD19+ IgM+** → B cells
- **CD56+ CD3−** → NK cells
- **CD14+ CD11c+** → classical monocytes
- **CD11c+ CD14−** → dendritic cells / non-classical monocytes

```{r annotate, eval=FALSE}
# Fill in after interpreting the heatmap
cluster_labels <- c(
  "1" = "CD4 T cells",
  "2" = "CD8 T cells",
  "3" = "NK cells",
  "4" = "B cells",
  "5" = "Monocytes"
  # ... add remaining clusters
)
expr_df$population <- cluster_labels[as.character(expr_df$cluster)]
```

```{r save}
saveRDS(expr_df, "data/processed/PBMC_clustered.rds")
```

<hr>
<div style="display:flex; gap:8px; margin-top:1.5em;">
<a href="05_dimensionality_reduction.html" class="btn btn-default btn-sm">← Prev: Dimensionality Reduction</a>
<a href="07_analysis.html" class="btn btn-primary btn-sm">Next: Analysis →</a>
</div>
