---
title: "06: Clustering (Optional)"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    includes:
      after_body: _nav.html
---

<div style="display:flex; gap:8px; margin-bottom:1.5em;">
<a href="05_dimensionality_reduction.html" class="btn btn-default btn-sm">← Prev: Dimensionality Reduction</a>
<a href="07_analysis.html" class="btn btn-primary btn-sm">Next: Analysis →</a>
</div>

---

```{r libraries, message=FALSE, warning=FALSE}
library(flowWorkspace)
library(FlowSOM)
library(tidyverse)
library(ggplot2)
```

```{r load}
gs      <- load_gs("data/processed/GvHD_tcell_gs")
umap_df <- readRDS("data/processed/GvHD_tcell_umap.rds")
```

This script is for discovery mode — when you want to find populations without defining them manually through gating. If your panel has predefined populations you already know how to gate, you may not need this step. Where clustering becomes useful is for high-dimensional panels where manual gating can't capture the full complexity (for example, identifying functional subsets within a gated population, or doing an unbiased survey of all live cells).

FlowSOM builds a self-organizing map on your expression data, then groups SOM nodes into metaclusters using hierarchical clustering. The two parameters you control are `xdim`/`ydim` (the SOM grid size, which determines resolution) and `maxMeta` (the maximum number of metaclusters to consider). Start conservatively and adjust — too few clusters merge populations that should be distinct, too many over-split into noise.

```{r extract_for_clustering}
fluor_channels <- c("CD15_FITC_H", "CD45_PE_H", "CD14_PerCP_H", "CD33_APC_H")

singlet_data <- gs_pop_get_data(gs, "singlets")

set.seed(42)
expr_list <- fsApply(singlet_data, function(ff) {
  idx <- sample(nrow(ff), min(500, nrow(ff)))
  df  <- as.data.frame(exprs(ff)[idx, fluor_channels])
  df$sample <- identifier(ff)
  df
}, simplify = FALSE)

expr_df <- bind_rows(expr_list)
```

```{r run_flowsom}
fsom <- FlowSOM(
  input   = as.matrix(expr_df[, fluor_channels]),
  xdim    = 7,
  ydim    = 7,
  maxMeta = 6,
  seed    = 42
)
```

Add cluster assignments back to the expression dataframe, then join with UMAP coordinates so we can see which UMAP regions correspond to which clusters.

```{r add_clusters, fig.width=6, fig.height=5, warning=FALSE}
expr_df$cluster <- as.factor(GetMetaclusters(fsom))

# The subsampled rows are in the same order as umap_df rows if seed was identical
umap_clustered <- umap_df %>%
  mutate(cluster = expr_df$cluster)

ggplot(umap_clustered, aes(x = UMAP1, y = UMAP2, color = cluster)) +
  geom_point(size = 0.8, alpha = 0.6) +
  theme_minimal() +
  labs(title = "FlowSOM metaclusters overlaid on UMAP")
```

To understand what each cluster represents, look at the median expression of each marker per cluster. Clusters with distinct marker combinations correspond to populations you'd recognize from gating or from biology.

```{r cluster_heatmap, fig.width=8, fig.height=4, warning=FALSE}
cluster_medians <- expr_df %>%
  group_by(cluster) %>%
  summarise(across(all_of(fluor_channels), median)) %>%
  pivot_longer(-cluster, names_to = "channel", values_to = "median_expression")

ggplot(cluster_medians, aes(x = channel, y = cluster, fill = median_expression)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Median marker expression per FlowSOM cluster",
       x = NULL, y = "Cluster")
```

Once you've identified what each cluster represents from the heatmap, annotate them. Some clusters may need to be split further or merged if they don't correspond to a biologically meaningful distinction.

```{r annotate, eval=FALSE}
# Fill in after interpreting the heatmap above
cluster_labels <- c(
  "1" = "Population A",
  "2" = "Population B",
  "3" = "Population C"
  # ... add remaining clusters
)

expr_df$population <- cluster_labels[as.character(expr_df$cluster)]
```

```{r save}
saveRDS(expr_df, "data/processed/GvHD_tcell_clustered.rds")
```

<hr>
<div style="display:flex; gap:8px; margin-top:1.5em;">
<a href="05_dimensionality_reduction.html" class="btn btn-default btn-sm">← Prev: Dimensionality Reduction</a>
<a href="07_analysis.html" class="btn btn-primary btn-sm">Next: Analysis →</a>
</div>
