```{r libraries}
library(flowWorkspace)
library(tidyverse)
library(ggplot2)
library(lme4)
library(emmeans)
```

```{r load}
gs <- load_gs("data/processed/20240603_CD8tet_gs")
meta <- pData(gs) %>% rownames_to_column("sample")
```

This script extracts the numbers we actually want to compare across groups: cell frequencies and marker expression (MFI). Everything before this was preparation — this is where we answer the biological question.

## Cell frequencies

We report frequencies as percentage of the parent gate rather than absolute cell counts. Absolute counts depend on how much blood or tissue you started with, which introduces variation unrelated to biology. Percentage of parent is the standard in flow cytometry for this reason.

```{r extract_frequencies}
# Get counts at each gating node
stats <- gs_pop_get_stats(gs, nodes = gs_get_pop_paths(gs), type = "percent")

# Focus on the population of interest as a percent of live cells
freq_df <- stats %>%
  filter(pop == "/lymphocytes/singlets/live/CD8_T_cells") %>%
  rename(sample = sample, pct_of_live = percent) %>%
  left_join(meta, by = "sample")

freq_df
```

Plot frequencies by treatment group. Use a dotplot with individual points rather than a bar chart — bar charts hide the sample-level distribution and make it hard to see outliers or small group sizes.

```{r plot_frequencies}
ggplot(freq_df, aes(x = treatment, y = pct_of_live, color = treatment)) +
  geom_jitter(width = 0.1, size = 2) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.3, color = "black") +
  theme_minimal() +
  labs(
    y = "CD8 T cells (% of live)",
    x = NULL,
    title = "CD8 T cell frequency by treatment"
  ) +
  theme(legend.position = "none")
```

## Marker expression (MFI)

MFI (median fluorescence intensity) tells you how much of a marker is expressed on a per-cell basis within a gated population. We use median rather than mean because the arcsinh-transformed distributions are sometimes skewed.

```{r extract_mfi}
cd8_data <- gs_pop_get_data(gs, "CD8_T_cells")
fluor_channels <- colnames(cd8_data[[1]])[4:ncol(cd8_data[[1]])]

mfi_df <- fsApply(cd8_data, function(ff) {
  apply(exprs(ff)[, fluor_channels], 2, median)
}, simplify = TRUE) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(meta, by = "sample")
```

```{r plot_mfi}
# Plot MFI for a specific marker across treatment groups
marker_channel <- "B515.A"  # replace with the channel for your marker of interest
marker_name <- "CD44"        # replace with the marker name

ggplot(mfi_df, aes(x = treatment, y = .data[[marker_channel]], color = treatment)) +
  geom_jitter(width = 0.1, size = 2) +
  stat_summary(fun = median, geom = "crossbar", width = 0.3, color = "black") +
  theme_minimal() +
  labs(
    y = paste(marker_name, "MFI (arcsinh)"),
    x = NULL,
    title = paste(marker_name, "expression on CD8 T cells")
  ) +
  theme(legend.position = "none")
```

## Statistical comparisons

We use a linear mixed model with cage as a random effect. Cage is nearly always a confound in mouse experiments — animals within the same cage share microbiome, bedding, and social environment. Including it as a random effect accounts for this without treating it as a fixed variable of interest.

```{r stats_frequency}
# Fit model
model_freq <- lmer(pct_of_live ~ treatment + (1 | cage), data = freq_df)
summary(model_freq)

# Pairwise comparisons between treatment groups
emmeans(model_freq, pairwise ~ treatment)
```

```{r stats_mfi}
model_mfi <- lmer(.data[[marker_channel]] ~ treatment + (1 | cage), data = mfi_df)
summary(model_mfi)

emmeans(model_mfi, pairwise ~ treatment)
```

```{r save_results}
dir.create("output/tables", recursive = TRUE, showWarnings = FALSE)

write_csv(freq_df, "output/tables/20240603_CD8tet_frequencies.csv")
write_csv(mfi_df, "output/tables/20240603_CD8tet_mfi.csv")
```
