---
title: "07: Statistical Analysis"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
---

<div style="display:flex; gap:8px; margin-bottom:1.5em; align-items:center; flex-wrap:wrap;">
<a href="../index.html" class="btn btn-default btn-sm">← Home</a>
<div class="btn-group">
<button type="button" class="btn btn-default btn-sm dropdown-toggle" data-toggle="dropdown">Scripts <span class="caret"></span></button>
<ul class="dropdown-menu">
<li><a href="01_load_and_metadata.html">01: Load and Metadata</a></li>
<li><a href="02_qc.html">02: QC</a></li>
<li><a href="03_transformation.html">03: Transformation</a></li>
<li><a href="04_gating.html">04: Gating</a></li>
<li><a href="05_dimensionality_reduction.html">05: Dimensionality Reduction</a></li>
<li><a href="06_clustering.html">06: Clustering</a></li>
<li><a href="07_analysis.html">07: Analysis</a></li>
</ul>
</div>
<a href="06_clustering.html" class="btn btn-default btn-sm">← Prev: Clustering</a>
</div>

---

```{r libraries, message=FALSE, warning=FALSE}
library(flowWorkspace)
library(tidyverse)
library(ggplot2)
library(lme4)
library(emmeans)
```

```{r load}
gs   <- load_gs("data/processed/GvHD_tcell_gs")
meta <- pData(gs) %>% rownames_to_column("sample")
```

This script extracts the numbers we actually want to compare across groups: cell frequencies and marker expression (MFI). Everything before this was preparation — this is where we answer the biological question.

## Cell frequencies

We report frequencies as percentage of the parent gate rather than absolute cell counts. Absolute counts depend on how much sample you started with, which introduces variation unrelated to biology. Percentage of parent is the standard in flow cytometry.

```{r extract_frequencies}
stats <- gs_pop_get_stats(gs, nodes = gs_get_pop_paths(gs), type = "percent")

freq_df <- stats %>%
  dplyr::filter(pop == "/lymphocytes/singlets/population_of_interest") %>%
  rename(pct_of_singlets = percent) %>%
  left_join(meta, by = "sample")

freq_df
```

Use a dotplot with individual points rather than a bar chart. Bar charts hide the sample-level distribution and make it hard to see outliers or small group sizes. The crossbar shows the group mean.

```{r plot_frequencies, fig.width=5, fig.height=4, warning=FALSE}
ggplot(freq_df, aes(x = treatment, y = pct_of_singlets, color = treatment)) +
  geom_jitter(width = 0.1, size = 2.5) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.3, color = "black") +
  theme_minimal() +
  labs(
    y     = "Population of interest (% of singlets)",
    x     = NULL,
    title = "Cell frequency by GvHD status"
  ) +
  theme(legend.position = "none")
```

Also worth plotting how frequency changes over time within patients:

```{r plot_longitudinal, fig.width=7, fig.height=4, warning=FALSE}
ggplot(freq_df, aes(x = timepoint_days, y = pct_of_singlets,
                    color = treatment, group = patient_id)) +
  geom_line(alpha = 0.6) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    x     = "Days post-transplant",
    y     = "Population of interest (% of singlets)",
    title = "Longitudinal cell frequency per patient"
  )
```

## Marker expression (MFI)

MFI (median fluorescence intensity) tells you how much of a marker is expressed on a per-cell basis within a gated population. We use median rather than mean because the distributions are sometimes skewed.

```{r extract_mfi}
fluor_channels <- c("CD15_FITC_H", "CD45_PE_H", "CD14_PerCP_H", "CD33_APC_H")

pop_data <- gs_pop_get_data(gs, "population_of_interest")

mfi_list <- fsApply(pop_data, function(ff) {
  vals <- apply(exprs(ff)[, fluor_channels, drop = FALSE], 2, median)
  as.data.frame(t(vals))
}, simplify = FALSE)

mfi_df <- bind_rows(mfi_list, .id = "sample") %>%
  left_join(meta, by = "sample")

head(mfi_df[, c("sample", "CD15_FITC_H", "CD45_PE_H", "treatment")])
```

```{r plot_mfi, fig.width=5, fig.height=4, warning=FALSE}
ggplot(mfi_df, aes(x = treatment, y = .data[["CD15_FITC_H"]], color = treatment)) +
  geom_jitter(width = 0.1, size = 2.5) +
  stat_summary(fun = median, geom = "crossbar", width = 0.3, color = "black") +
  theme_minimal() +
  labs(
    y     = "CD15 MFI",
    x     = NULL,
    title = "CD15 expression by GvHD status"
  ) +
  theme(legend.position = "none")
```

## Statistical comparisons

This is a longitudinal dataset with repeated measurements per patient — the same patients are measured at multiple visits. We use a linear mixed model with patient as a random effect to account for this. For mouse experiments, use cage as the random effect instead (see model comment below).

```{r stats_frequency}
model_freq <- lmer(pct_of_singlets ~ treatment + timepoint_days + (1 | patient_id),
                   data = freq_df)
summary(model_freq)
```

```{r emmeans_frequency}
emmeans(model_freq, pairwise ~ treatment)
```

```{r stats_mfi}
model_mfi <- lmer(mfi_df[["CD15_FITC_H"]] ~ treatment + timepoint_days + (1 | patient_id),
                  data = mfi_df)

emmeans(model_mfi, pairwise ~ treatment)
```

> For mouse experiments, replace `(1 | patient_id)` with `(1 | cage)` and `timepoint_days` with your experimental timepoint variable. Always include cage as a random effect in mouse studies — animals within the same cage share environment, microbiome, and social conditions, and this is almost always a confound.

```{r save_results}
dir.create("output/tables", recursive = TRUE, showWarnings = FALSE)

write_csv(freq_df, "output/tables/GvHD_tcell_frequencies.csv")
write_csv(mfi_df,  "output/tables/GvHD_tcell_mfi.csv")
```

<hr>
<div style="display:flex; gap:8px; margin-top:1.5em; align-items:center; flex-wrap:wrap;">
<a href="../index.html" class="btn btn-default btn-sm">← Home</a>
<div class="btn-group">
<button type="button" class="btn btn-default btn-sm dropdown-toggle" data-toggle="dropdown">Scripts <span class="caret"></span></button>
<ul class="dropdown-menu">
<li><a href="01_load_and_metadata.html">01: Load and Metadata</a></li>
<li><a href="02_qc.html">02: QC</a></li>
<li><a href="03_transformation.html">03: Transformation</a></li>
<li><a href="04_gating.html">04: Gating</a></li>
<li><a href="05_dimensionality_reduction.html">05: Dimensionality Reduction</a></li>
<li><a href="06_clustering.html">06: Clustering</a></li>
<li><a href="07_analysis.html">07: Analysis</a></li>
</ul>
</div>
<a href="06_clustering.html" class="btn btn-default btn-sm">← Prev: Clustering</a>
</div>
